## Reproducible Research: Peer Assessment 1

### Initial Preparation
Before we get started on answering the questions, make sure of the following:

- Fork the RepData_PeerAssessment1 directory at github
- Use the `setwd("path/where/the/files/are/located")`
- Unzip the activity.zip file to obtain activty.csv file


### Loading and preprocessing the data

Load the data into R:

```{r loadData, echo=TRUE}
activity <- read.csv("activity.csv", header=T)
```

To see what kind of variables are listed in the activity data frame

```{r, echo=TRUE}
str(activity)
```

The variables are steps, date and interval. Note that date variable is a factor. Convert the date variable from a factor into a proper R date variable. 

```{r, echo=TRUE}
activity[,2] <- as.Date(activity[,2]) # The second column is the date variable
class(activity$date)                  # To ensure that the date variable is the date format
```

### What is mean total number of steps taken per day?

Sum the values by date using the `tapply()` function. We convert the resulting array into a data frame and use the ggplot2 package to get the histogram of the total no of steps.

```{r, echo=TRUE, fig.width=6, fig.height=6}
totalSteps <- tapply(activity$steps, activity$date, sum)     # Returns an array of the sums by date
totalStepsDf  <- as.data.frame(totalSteps)                   # Convert array to a data frame
totalStepsDf$date <- rownames(totalStepsDf)                  # Convert the rownames into a column
rownames(totalStepsDf) <- NULL                               # Remove the rownames
library(ggplot2)                                             # Using the ggplot2 package
hist <- qplot(totalSteps, data = totalStepsDf)               # Set the plot
hist + geom_histogram(binwidth=2000) +                       # Set the width of the bins
  labs(y = "Count") +                                        # Set label for y axis
  labs(x = "Total No of Steps") +                            # Set label for x axis
  labs(title = "Total No of Steps per day")                  # Set the title of the plot

## Calculate the mean and median
meanSteps  <- mean(totalStepsDf$totalSteps, na.rm=T)         # Mean no of steps
medianSteps <- median(totalStepsDf$totalSteps, na.rm = T)    # Median no of steps
```

The mean no of total steps is `r meanSteps` and the median no of steps is `r medianSteps`.

### What is the average daily activity pattern?

Use the `tapply()` function to calculate the mean by the interval

```{r, echo=TRUE, fig.width=6, fig.height=6}
meanByInterval <- tapply(activity$steps, activity$interval,
    function(m) mean(m, na.rm=T)
)
plot(meanByInterval, type="l",                              # Plot the line graph
         xlab="Interval",                                   
         ylab="No of steps",
         main="Mean no of Steps"
     )

maxSteps  <- max(meanByInterval)
maxMean <- max(meanByInterval)
maxSteps <- meanByInterval[meanByInterval == maxMean]
```

The 5-minute interval with the maximum number of steps is `r names(maxSteps)[1]`.

### Imputing missing values

Code for imputing the valuee:

```{r,echo=TRUE}
totalNA <- sum(!complete.cases(activity))

## Impute the NA values with the mean and create a new dataset
newActivity <- activity
for (i in 1:nrow(activity)) {
  if (is.na(activity$steps[i])) {
    newActivity$steps[i] <- meanByInterval[as.character(activity$interval[i])]
  }
}

## Create a histogram with the new dataset
newtotalSteps <- tapply(newActivity$steps, newActivity$date, sum)
newtotalStepsDf  <- as.data.frame(newtotalSteps)
newtotalStepsDf$date <- rownames(newtotalStepsDf)
rownames(newtotalStepsDf) <- NULL
library(ggplot2) 
newhist <- qplot(totalSteps, data = newtotalStepsDf)
newhist + geom_histogram(binwidth=2000) + 
  labs(y = "Count") + 
  labs(x = "Total No of Steps") + 
  labs(title = "Total No of Steps per day (Imputed NAs)")

## Calculate the mean and median for the new dataset
newmeanSteps  <- mean(newtotalStepsDf$newtotalSteps)
newmedianSteps <- median(newtotalStepsDf$newtotalSteps)

## Comparison between the values
table <- data.frame(mean=c(meanSteps,newmeanSteps), median=c(medianSteps,newmedianSteps))
rownames(table) <- c("withNA", "new")
print(table)

boxplot(totalSteps, newtotalSteps, names=c("withNAs","new"))
```

Total no of missing values is `r totalNA`.

For the new dataset with the imputed NA values the mean total no of steps is `r newmeanSteps` and the median total no of steps is `r newmedianSteps`.

There is no differnce in the means, however the median is higher for the new dataset. Imputing the NAs in the dataset did not have a resonable impact on the mean and median values. From the boxplots it is clear that the range and the interquartile range is narrower for the new dataset.


### Are there differences in activity patterns between weekdays and weekends?

Use the `weekdays()` function to convert the date into a factor. We are using the newActivity dataset (has no NA values)

```{r, echo=TRUE}
day <- weekdays(newActivity$date, abbreviate=T)

## Create a new column day and two levels, weekends and weekdays
weekends  <- c("Sat", "Sun")
weekdays <- c("Mon","Tue","Wed","Thu","Fri")
newActivity$day[day %in% weekends] <- "weekend"
newActivity$day[day %in% weekdays] <- "weekday"
newActivity$day <- factor(newActivity$day)

# Split data by the day column
splitByDay <- split(newActivity, newActivity$day)
weekdayActivity  <- splitByDay$weekday
weekendActivity <- splitByDay$weekend

meanByIntervalWeekdays <- tapply(weekdayActivity$steps,
                                    weekdayActivity$interval,
                                    mean
                                )
meanByIntervalWeekends <- tapply(weekendActivity$steps,
                                    weekendActivity$interval,
                                    mean
                                )

## Plot the panel graphs
par(mfrow=c(2,1))
plot(meanByIntervalWeekdays, type="l",
         main="Weekday",
         xlab="Interval",
         ylab="No of Steps"
     )
plot(meanByIntervalWeekends, type="l",
         main="Weekend",
         xlab="Interval",
         ylab="No of Steps"
     )
```

